Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-02 appendix.tex
A few notes.

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
%das: went with "conditional likelihood"
32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the conditional likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
Assume $x_1,y_1,x_2 < 1$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

@@

Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
We now prove the main result.

@@

%EM Where is this shown in the previous proof?
%das: wrote out more explicitly. It's similar to the argument starting at L612.

@@

The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
%EM Hm... it seems like we need to exclude having the b's very close to 1. How do we do that?
%das: I thought that was taken care of by the condition above: $b_{2}^U, b_{1}^U, b_{12}^U < 1$. It should also be taken care of by the below inequality, too, right? Otherwise as b -> 1 this will go to -\infty, so the below will fail. Do you think they need to be well-separated from 1 instead of just strictly less than 1?
\ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}

>>>>

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the conditional likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
%EM We don't think they occur.
Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Perhaps we can say something about how this is not a strong assumption.
%EM see below "By the assumption of"
Assume $x_1,y_1,x_2 < 1$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

@@

%EM Also here.
Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
We now prove the main result.

@@

@@

%EM Topic sentence here please.
The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
\[
\ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}
